The Galactic Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-16 paper/ms.tex
Update abstract

\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{bm}
\usepackage{mathtools}

@@

\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\avg}[1]{\left< #1 \right>}
\newcommand{\z}{z_r}
\newcommand{\uth}{\textsuperscript{th}}
\newcommand{\n}{\text{n}}

@@

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. Of course, inaccurate measurements of the 
phase space position and uncertainities in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as an error. This error is
non-Gaussian and bimodal, 
%mm [this sounds imprecise: an error wouldn't peak on a value...]
with neither of the modes peaking on the true action value. 
A midplane error of $\sim30\,\pc$ for a thin disk orbit or
$\sim225\,\pc$ for a thick disk orbit is all that is necessary to
induce a $100\,\%$ error in the vertical action $J_z$.
orbits, respectively. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic offset in $z$. Variation in the local standard of rest
will induce similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across large regions of the disk
accessible with the {\em Gaia} mission.

@@

\subsection{Epicyclic Approximation} \label{ssec:epi_action}
Before turning to numerical methods, we first write down analytically the
induced action error from either a midplane offset or a Galactocentric radius
offset. We make the epicyclic approximation, which assumes that the motion in
the $z$ and $R$ components of the orbit are decoupled and follow a simple
harmonic oscillator. This approximation should be an excellent description of
thin disk-like orbits and a good description of thick disk-like orbits. We
also make the assumption of a perfectly flat circular velocity curve with
velocity $v_c$.

Under this approximation, we can write down the vertical and radial components
of the orbits as,
\beq\label{eq:orbits_epi}
\begin{split}
z(t) &= A_z \sin{(\kappa t + \alpha)}\\
R(t) &= A_R \sin{(\nu t + \beta)}\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the vertical and radial (or epicyclic)
frequencies, $A_z$ and $A_R$ are the amplitudes of the stellar motion in the
vertical and radial coordinates, and $\alpha$ and $\beta$ are arbitrary
constants. We can write down the actions in this approximation as
\citep[Section~3.5.3b][]{2008gady.book.....B},
\beq\label{eq:actions_epi}
\begin{split}
L_z &= R_g v_c \\
J_r &= \kappa A_R^2 \\
J_z &= \nu A_z^2\text{,}
\end{split}
\eeq
where $R_g$ is the guiding radius of the orbit.

An induced error in $z$ of $\Delta z$ will cause $\Delta A_z \sim \Delta z$.
Similarly, an induced error in $R$ will cause $\Delta A_R \sim \Delta R$. If
we assume that $\kappa$ and $\nu$ are constant (\Gus{justify?}), then we have
that
\beq\label{eq:actions_epi_error}
\begin{split}
\Delta L_z = \Delta R v_c &\implies \frac{\Delta L_z}{L_z} = \frac{\Delta R}{R} \\
\Delta J_r = 2 \kappa \Delta A_R &\implies \frac{\Delta J_r}{J_r} = \frac{2\Delta R}{A_R} \\
\Delta J_z = 2 \nu \Delta A_z &\implies \frac{\Delta J_z}{J_z} = \frac{2\Delta z}{A_z}\text{.}
\end{split}
\eeq
Therefore, errors in each action are linearly proportional to the errors in
either the midplane or cylindrical radius.

There is a slight complication in the formula for computing $\Delta J_r/J_r$.
In a moment we will consider the effect of an inaccuracy in the distance from
the Sun to the Galactic center on actions. However, this inaccuracy induces an
error in $x$, or $\Delta x$. This is not the same as an inaccuracy in $\Delta
R$, but they are related by the following argument. An error $\Delta x$ will
induce an erroneous radius $R_{\text{err}}$ related by the formula,
\beq
(x+\Delta x)^2 + y^2 = R_{\text{err}}^2\text{.}
\eeq
Keeping only terms to first order in $\Delta x$, we have that,
\beq
\begin{split}
R_{\text{err}}^2 &= R^2 - 2 R \cos{\theta} \Delta x \\
\implies \Delta R &\equiv \abs{R_{\text{err}} - R} = \abs{\cos{\theta}} \Delta x\text{.}
\end{split}
\eeq
Averaging over the circle, we therefore have that
\beq
\avg{\Delta R} = \frac{2}{\pi} \Delta x\text{.}
\eeq
Therefore, the error in $J_r$ is related to $\Delta x$ by,
\beq\label{eq:Jr_epi_dx}
\frac{\Delta J_r}{J_r} = \frac{4 \Delta x}{\pi A_R}\text{.}
\eeq

We will check these formulae against numerical tests next.

>>>>

\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{bm}

@@

\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\z}{z_r}
\newcommand{\uth}{\textsuperscript{th}}
\newcommand{\n}{\text{n}}

@@

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. Of course, inaccurate measurements of the 
phase space position and uncertainities in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as an error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on the true action value. 
A midplane error of $\sim30\,\pc$ for a thin disk orbit or
$\sim225\,\pc$ for a thick disk orbit is all that is necessary to
induce a $100\,\%$ error in the vertical action $J_z$.
orbits, respectively. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic offset in $z$. Variation in the local standard of rest
will induce similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across large regions of the disk
accessible with the {\em Gaia} mission.

@@
